\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-29 manuscript/Rmd/draft.Rmd
3.2 added intro
Update Rmd

The first sets of data that we rely on are historical results from major climbing competitions in the past. We collected data on climbing contests that took place between 2018 and 2020, where the combined format was used to determine the scores and ranks of climbers. The events include the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Youth Olympics. Data were obtained from various sources, including the event websites, Wikipedia, and the International Federation of Sport Climbing (IFSC). The main attributes of our datasets are the name and nationality of the climbers; bib number (for some competitions); the finishing place of climbers in speed climbing, bouldering, and lead climbing; the total score (which equals the product of event ranks); and the final rank. We utilize this data to compute the correlations between the event ranks and final table position, as well as to look at how often the final orderings change if one athlete is dropped and the ranks for each discipline are re-computed.

Our second data source comes from a simulation study that we conducted, with the purpose of examining the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the total scores for every simulated round, as well as the final standings for the climbing athletes. The simulation results allow to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions.

@@

```{r}
library(tidyverse)
library(knitr)
library(xtable)
theme_set(theme_bw())
```

@@

@@

For the qualification round, a climber is almost guaranteed to make the final round if they win the first event (with a 99.51% chance of advancing) or if they win at least one of the three climbing concentrations (99.48%). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 1. We notice that on average, the minimum score that one should aim for in order to move on to the final round is 435 (for 8th rank).

```{r}
qual %>% 
  group_by(rank) %>% 
  summarize(avg_adv_score = mean(score)) %>% 
  filter(rank <= 8) %>% 
  xtable(caption = "Average score for each qualifying rank") %>% 
  print()
```

Regarding the finals, a climber is very likely to earn a medal (finish in the top 3) if they win the first event (83.03% chance) or any event (85.01%). In order to obtain a climbing medal, the average score for getting gold, silver, and bronze are 9.6748, 20.4143, and 33.2648, respectively (Table 2).

```{r}
final %>% 
  group_by(rank) %>% 
  summarize(avg_score = mean(score)) %>% 
  xtable(caption = "Average score of medalists") %>% 
  print()
```

```{r, fig.cap = "Boxplots of scoring distribution for every qualification rank", eval = FALSE}
qual %>% 
  mutate(rank = factor(rank)) %>% 
  ggplot(aes(x = score, y = rank)) +
  geom_boxplot()
```

@@

<!-- ### Sum of ranks instead of product -->

<!-- We performed the same tasks as we did for products -->

<!-- Probabilities of advancing, winning medals are all lower -->

<!-- Most notably for qualifications,  -->

<!-- Obvious that the average score between the ranks are closer to each other -->

<!-- The amount of variability doesn't seem to be different as the rank increases -->

## Correlations

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 3 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau (Kendall Rank Correlation Coefficient) as our measure of ordinal association between the quantities. It is evidently clear that there is a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations.

This trend also holds for most of the past competitions.

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Women's Qualification"}
wq %>% 
  select(rank, speed, bould, lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank())
```

@@

@@

The cases where someone behind you drops out and your ranking changes.

Example from 2018 youth, women's final
dropping ranks 3 and 5 change the medalist order

@@

```{r}
mf_gg <- drop_rerank(mf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 4) | 
             rank_drop == 1 & rank %in% 4:5 |
              rank_drop == 2 & rank %in% c(1, 5) |
                rank_drop %in% c(3, 5) & rank %in% c(1, 2, 5) |
                  rank_drop == 6 & rank %in% 1:3, 
           "no", "yes")) %>% 
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(1, 2, 3, 5, 6)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2) +
  expand_limits(y = 102, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(x = "Men", y = "") +
  theme_minimal() +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())

wf_gg <- drop_rerank(wf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 1, 4, 6) |
             rank_drop == 2 & rank %in% 1:3 |
              rank_drop == 3 & rank %in% c(1, 4, 5) |
                rank_drop == 5 & rank %in% c(1, 5),
           "no", "yes")) %>%
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(2, 3, 5)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2, ncol = 4) +
  expand_limits(y = 62, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(y = "\nScore", x = "Women") +
  theme_minimal() +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

>>>>

Our first sets of data come from a simulation study that we conducted, with the purpose of examining the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the total scores for every simulated round, as well as the final standings for the climbing athletes. The simulation results allow to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions.

@@

```{r}
library(tidyverse)
library(knitr)
library(xtable)
theme_set(theme_minimal())
```

@@

Additionally, we rely historical results from major climbing competitions in recent years. We collected data on climbing contests that took place between 2018 and 2020, where the combined format was used to determine the scores and ranks of climbers. The events include the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Youth Olympics. Data were obtained from various sources, including the event websites, Wikipedia, and the International Federation of Sport Climbing (IFSC). The main attributes of our datasets are the name and nationality of the climbers; bib number (for some competitions); the finishing place of climbers in speed climbing, bouldering, and lead climbing; the total score (which equals the product of event ranks); and the final rank. We utilize this data to compute the correlations between the event ranks and final table position, as well as to look at how often the final orderings change if one athlete is dropped and the ranks for each discipline are re-computed.

@@

In this section, we discuss the results of our simulations described in Section \ref{data-and-methods}. For the qualification round, our simulation study shows that a climber is almost guaranteed to make the final round if they win the first event (with a 99.51% chance of advancing) or if they win at least one of the three climbing concentrations (99.48%). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 2. We notice that on average, the minimum score that one should aim for in order to move on to the final round is about 435 (for 8th rank).

```{r}
qual %>% 
  rename(Rank = rank) %>% 
  group_by(Rank) %>% 
  summarize(`Average score` = mean(score)) %>% 
  filter(Rank <= 10) %>% 
  xtable(caption = "The average scores for the top 10 qualification ranks according to our simulations. A climber will secure a finalist spot if they finish in the top 8.") %>% 
  print()
```

Regarding the finals, a climber is very likely to finish in the top 3 and hence earn a medal if they win the first event (83.03% chance) or any event (85.01% chance). Furthermore, according to our final simulations, in order to obtain a climbing medal, the average scores that put an athlete in position to receive gold, silver, and bronze medals are 9.6748, 20.4143, and 33.2648, respectively (see Table 3).

```{r}
final %>% 
  rename(Rank = rank) %>% 
  group_by(Rank) %>% 
  summarize(`Average score` = mean(score)) %>% 
  xtable(caption = "The average scores for all final ranks according to our simulations. Ranks 1, 2, and 3 are table positions that guarantee medalist status for climbers.") %>% 
  print()
```

@@

## Rank Correlation

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 2 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau (Kendall Rank Correlation Coefficient) as our measure of ordinal association between the quantities. It is evidently clear that there exists a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations. 
<!-- This trend also holds for most of the past competitions. -->

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Women's Qualification"}
wq %>% 
  select(rank, speed, bould, lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme_bw() +
  theme(panel.grid.major = element_blank())
```

@@

Another interesting question that we are interested in investigating is "What would happen to the rankings if a single climber is removed?".

This illustrates the idea of IIA

We once again make use of data from the 2018 Youth Olympics for this analysis, but this time we examine the final round of both men's and women's competitions.

Figure 3 shows the plots

@@

The cases where someone behind you drops out and your ranking changes. Figure

@@

```{r}
mf_gg <- drop_rerank(mf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 4) | 
             rank_drop == 1 & rank %in% 4:5 |
              rank_drop == 2 & rank %in% c(1, 5) |
                rank_drop %in% c(3, 5) & rank %in% c(1, 2, 5) |
                  rank_drop == 6 & rank %in% 1:3, 
           "no", "yes")) %>% 
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(1, 2, 3, 5, 6)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2) +
  expand_limits(y = 102, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(x = "Men", y = "") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())

wf_gg <- drop_rerank(wf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 1, 4, 6) |
             rank_drop == 2 & rank %in% 1:3 |
              rank_drop == 3 & rank %in% c(1, 4, 5) |
                rank_drop == 5 & rank %in% c(1, 5),
           "no", "yes")) %>%
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(2, 3, 5)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2, ncol = 4) +
  expand_limits(y = 62, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(y = "\nScore", x = "Women") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

### 2021-07-29 manuscript/Rmd/draft.tex
Update Rmd

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\usepackage{caption}
\captionsetup[table]{skip=3pt}
\usepackage{float}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Histogram of simulated scores for qualification and final
rounds}
\end{figure}

@@

For the qualification round, a climber is almost guaranteed to make the
final round if they win the first event (with a 99.51\% chance of
advancing) or if they win at least one of the three climbing
concentrations (99.48\%). On the other hand, finishing last in the first
event or in any event would certainly hurt an athlete's chance of
finishing in the top 8, as the probabilities of a climber advancing
given they finish last in the first and in any event are 0.1830 and
0.1885, respectively. In addition, the average score for qualification
positions 1 to 8 are displayed in Table 1. We notice that on average,
the minimum score that one should aim for in order to move on to the
final round is 435 (for 8th rank).

\begin{table}[ht]
\centering
\caption{Average score for each qualifying rank} 
\begin{tabular}{rr}
  \hline
rank & avg\_adv\_score \\ 
  \hline
  1 & 36.02 \\ 
    2 & 73.61 \\ 
    3 & 115.40 \\ 
    4 & 162.23 \\ 
    5 & 216.00 \\ 
    6 & 278.16 \\ 
    7 & 350.33 \\ 
    8 & 434.59 \\ 
   \hline
\end{tabular}
\end{table}

Regarding the finals, a climber is very likely to earn a medal (finish
in the top 3) if they win the first event (83.03\% chance) or any event
(85.01\%). In order to obtain a climbing medal, the average score for
getting gold, silver, and bronze are 9.6748, 20.4143, and 33.2648,
respectively (Table 2).

\begin{table}[ht]
\centering
\caption{Average score of medalists} 
\begin{tabular}{rr}
  \hline
rank & avg\_score \\ 
  \hline
  1 & 9.67 \\ 
    2 & 20.41 \\ 
    3 & 33.26 \\ 
    4 & 50.59 \\ 
    5 & 74.76 \\ 
    6 & 110.05 \\ 
    7 & 164.43 \\ 
    8 & 265.78 \\ 
   \hline
\end{tabular}
\end{table}

@@

\hypertarget{correlations}{%
\subsection{Correlations}\label{correlations}}

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 3 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau (Kendall Rank Correlation Coefficient) as our measure of
ordinal association between the quantities. It is evidently clear that
there is a strong and positive correlation between the ranks of
bouldering and lead climbing, and as a results, the standings of these
two events are highly correlated with the final rankings. On the other
hand, the correlation with the final rank is not as strong for speed
climbing. Thus, speed climbers are facing a huge disadvantage in this
scoring system, compared to those that are specialized in the other two
concentrations.

This trend also holds for most of the past competitions.

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Women's
Qualification}
\end{figure}

@@

@@

The cases where someone behind you drops out and your ranking changes.
Example from 2018 youth, women's final dropping ranks 3 and 5 change the
medalist order

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-13-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out a
single climber. For each gender competition, each panel represents the
rank of the drop-out athlete, with 0 being the original final results.
Each case with a change in rank orderings is highlighted by a black
panel border, and any player with a rank change is represented by a
red-filled bar}
\end{figure}

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Histogram of simulated scores for qualification and final
rounds}
\end{figure}

Additionally, we rely historical results from major climbing
competitions in recent years. We collected data on climbing contests
that took place between 2018 and 2020, where the combined format was
used to determine the scores and ranks of climbers. The events include
the 2020 Continental Championships of Europe, Africa, Oceania,
Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and
2018 Youth Olympics. Data were obtained from various sources, including
the event websites, Wikipedia, and the International Federation of Sport
Climbing (IFSC). The main attributes of our datasets are the name and
nationality of the climbers; bib number (for some competitions); the
finishing place of climbers in speed climbing, bouldering, and lead
climbing; the total score (which equals the product of event ranks); and
the final rank. We utilize this data to compute the correlations between
the event ranks and final table position, as well as to look at how
often the final orderings change if one athlete is dropped and the ranks
for each discipline are re-computed.

@@

In this section, we discuss the results of our simulations described in
Section \ref{data-and-methods}. For the qualification round, our
simulation study shows that a climber is almost guaranteed to make the
final round if they win the first event (with a 99.51\% chance of
advancing) or if they win at least one of the three climbing
concentrations (99.48\%). On the other hand, finishing last in the first
event or in any event would certainly hurt an athlete's chance of
finishing in the top 8, as the probabilities of a climber advancing
given they finish last in the first and in any event are 0.1830 and
0.1885, respectively. In addition, the average score for qualification
positions 1 to 8 are displayed in Table 2. We notice that on average,
the minimum score that one should aim for in order to move on to the
final round is about 435 (for 8th rank).

\begin{table}[ht]
\centering
\caption{The average scores for the top 10 qualification ranks according to our simulations. A climber will secure a finalist spot if they finish in the top 8.} 
\begin{tabular}{rr}
  \hline
Rank & Average score \\ 
  \hline
  1 & 36.02 \\ 
    2 & 73.61 \\ 
    3 & 115.40 \\ 
    4 & 162.23 \\ 
    5 & 216.00 \\ 
    6 & 278.16 \\ 
    7 & 350.33 \\ 
    8 & 434.59 \\ 
    9 & 532.14 \\ 
   10 & 642.33 \\ 
   \hline
\end{tabular}
\end{table}

Regarding the finals, a climber is very likely to finish in the top 3
and hence earn a medal if they win the first event (83.03\% chance) or
any event (85.01\% chance). Furthermore, according to our final
simulations, in order to obtain a climbing medal, the average scores
that put an athlete in position to receive gold, silver, and bronze
medals are 9.6748, 20.4143, and 33.2648, respectively (see Table 3).

\begin{table}[ht]
\centering
\caption{The average scores for all final ranks according to our simulations. Ranks 1, 2, and 3 are table positions that guarantee medalist status for climbers.} 
\begin{tabular}{rr}
  \hline
Rank & Average score \\ 
  \hline
  1 & 9.67 \\ 
    2 & 20.41 \\ 
    3 & 33.26 \\ 
    4 & 50.59 \\ 
    5 & 74.76 \\ 
    6 & 110.05 \\ 
    7 & 164.43 \\ 
    8 & 265.78 \\ 
   \hline
\end{tabular}
\end{table}

@@

\hypertarget{rank-correlation}{%
\subsection{Rank Correlation}\label{rank-correlation}}

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 2 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau (Kendall Rank Correlation Coefficient) as our measure of
ordinal association between the quantities. It is evidently clear that
there exists a strong and positive correlation between the ranks of
bouldering and lead climbing, and as a results, the standings of these
two events are highly correlated with the final rankings. On the other
hand, the correlation with the final rank is not as strong for speed
climbing. Thus, speed climbers are facing a huge disadvantage in this
scoring system, compared to those that are specialized in the other two
concentrations.

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Women's
Qualification}
\end{figure}

@@

Another interesting question that we are interested in investigating is
``What would happen to the rankings if a single climber is removed?''.

This illustrates the idea of IIA

We once again make use of data from the 2018 Youth Olympics for this
analysis, but this time we examine the final round of both men's and
women's competitions.

Figure 3 shows the plots

@@

The cases where someone behind you drops out and your ranking changes.
Figure

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out a
single climber. For each gender competition, each panel represents the
rank of the drop-out athlete, with 0 being the original final results.
Each case with a change in rank orderings is highlighted by a black
panel border, and any player with a rank change is represented by a
red-filled bar}
\end{figure}
